The ``Dark Side'' of Phylogenetic Comparative Methods$^\dag$

### 2015-06-08 Manuscript/dark.tex
Minor edits to intro section

\title{The ``Dark Side'' of Phylogenetic Comparative Methods$^\dag$}
\author{
  Natalie Cooper$^{1,2,*}$ and Richard G. FitzJohn$^{4}$ 
  %Cecile An\'{e}$^{2}$, Wayne Maddison$^{3}$,\\  Gavin H. Thomas$^{5}$, April Wright$^{6}$,\\
}
\date{}
\affiliation{\noindent{\footnotesize

  $^1$ School of Natural Sciences, Trinity College Dublin, Dublin 2, Ireland.\\ 
  $^2$ Trinity Centre for Biodiversity Research, Trinity College Dublin, Dublin 2, Ireland.\\
  $^3$ Department of Animal and Plant Sciences, University of Sheffield, Sheffield S10 2TN, UK.\\
  $^4$ Department of Biological Sciences, Macquarie University, Sydney, NSW 2109, Australia. \\
  $^*$ Corresponding author: ncooper@tcd.ie; Zoology Building, Trinity College Dublin, Dublin 2, Ireland. 
       Fax: +353 1 677 8094; Tel: +353 1 896 1926.\\
  $^\dag$This paper is based on a Society of Systematic Biologists symposium at Evolution 2014, in Raleigh NC.\\
}}

\vfill
%\paragraph{Word-count:} X words

@@

\noindent
A long time ago in a galaxy far, far away - also known as the 1980s - phylogenetic comparative methods (PCMs) were first developed. 
These methods use information on the evolutionary relationships of organisms to compare species \citep{harvey1991comparative}, and were initially designed to deal with the statistical non-independence of species in comparative analyses (e.g. \citealp{felsenstein1985phylogenies,grafen1989phylogenetic}).  
% RF: I think we need a definition of PCMs here, around the second sentence?
% NC: I've had a go, see what you think...
Since then PCMs have been extended to investigate evolutionary pattern and process, and include methods for investigating drivers of diversification, the tempo and mode of trait evolution, and models of speciation and extinction (see reviews in \citealp{o2012evolutionary, pennell2013integrative}). 
PCMs have also become extremely popular over recent years; the number of papers containing the phrase ``phylogenetic comparative'' has increased exponentially since the 1980s (Figure \ref{PCMCitations}). 
With new methods being published almost weekly, there has never been a better time to be a comparative biologist.

@@

One major issue is that two groups of people are involved with PCMs. 
At one extreme, methods-developers develop new methods and implement them. They are (generally) aware of the limitations of their methods, and the assumptions that underly them. 
At the other extreme, na\"{i}ve methods-users just want to use PCMs to answer empirical questions with their data and, although they have expertise in other biological domains, they may not be aware of issues with the methods (of course the majority of people fall somewhere between these two extremes on this spectrum; many methods-developers are developing methods with a specific empirical question in mind, while there are many advanced methods-users who have detailed knowledge of the limitations of the methods they use). 
The problem is that although methods-developers (and advanced methods-users) are aware of the problems with PCMs, this information is not being effectively transferred to methods-users. 
Additionally, the tools and approaches used to fit models are often far more user-friendly and better documented than the methods used to to assess whether that model fit is reasonable. 
Clearly more effort is needed to bridge the widening gap between methods-users and methods-developers. Here we explore the causes of this communication gap and suggest some potential solutions. % NC: still not happy with this final line or two.

@@

\section{Acknowledgments}
Thanks to the Society of Systematic Biologists for funding our symposium at Evolution 2014. 
We also thank George Lucas for the ``hilarious'' Star Wars jokes.
%NC: Should possibly thank Brian O'Meara for funding discourse page

>>>>

\title{The ``Dark Side'' of Phylogenetic Comparative Methods$^\dag$}
\author{
  Natalie Cooper$^{1,2,*}$, Gavin H. Thomas$^{3}$\\ and Richard G. FitzJohn$^{4}$
}
\date{}
\affiliation{\noindent{\footnotesize
  $^1$ School of Natural Sciences, Trinity College Dublin, Dublin 2, Ireland.\\ 
  $^2$ Department of Life Sciences, Natural History Museum, Cromwell Road, London, SW7 5BD, UK.\\
  $^3$ Department of Animal and Plant Sciences, University of Sheffield, Sheffield, S10 2TN, UK.\\
  $^4$ Department of Biological Sciences, Macquarie University, Sydney, NSW 2109, Australia. \\
  $^*$ Corresponding author: nhcooper12@gmail.com; Department of Life Sciences, Natural History Museum, Cromwell Road, London, SW7 5BD, UK. Fax: +353 1 677 8094; Tel: +353 1 896 1926.\\
  $^\dag$This paper is based on a Society of Systematic Biologists symposium at Evolution 2014, in Raleigh NC.\\
}}

\vfill

@@

\noindent
A long time ago in a galaxy far, far away - also known as the 1980s - phylogenetic comparative methods (PCMs) were first developed. 
These methods use information on the evolutionary relationships of organisms to compare species \citep{harvey1991comparative}, and were initially designed to deal with the statistical non-independence of species in comparative analyses (e.g. \citealp{felsenstein1985phylogenies,grafen1989phylogenetic}).  
Since then PCMs have been extended to investigate evolutionary pattern and process, and include methods for investigating drivers of diversification, the tempo and mode of trait evolution, and models of speciation and extinction (see reviews in \citealp{o2012evolutionary, pennell2013integrative}). 
PCMs have also become extremely popular over recent years; the number of papers containing the phrase ``phylogenetic comparative'' has increased dramatically since the 1980s (Figure \ref{PCMCitations}). 
With new methods being published almost weekly, there has never been a better time to be a comparative biologist.

@@

We believe that a major cause of this problem is that although people developing and implementing new PCMs are aware of the limitations of these methods and the assumptions that underly them, this information is not being effectively transferred to methods-users.
Additionally, the tools and approaches used to fit models are often far more user-friendly and better documented than the methods used to to assess whether that model fit is reasonable. 
Clearly more effort is needed to bridge the widening gap between methods-users and methods-developers. 
Here we explore the causes of this communication gap and suggest some potential solutions.

@@

\section{Acknowledgments}
Thanks to the Society of Systematic Biologists for funding our symposium at Evolution 2014. 
%NC: Should possibly thank Brian O'Meara for funding discourse page
